Phylogenetics and the human microbiome.

### 2014-04-11 reviews/response.txt

> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
> believe it is meant to be a hybrid approach (BLAST being
> compositional, but the taxonomic labeling considered phylogenetic),
> but this could be clarified.

> 13-29. It is not clear what heuristics are required for a cut-off
> method. The author says that he is going to describe the OTU inference
> industry, but the paragraph is really just a list of acronyms, and
> then refers obliquely to a "contingency table" of OTU observations.
> So, for example, what is a notion of goodness of clustering that the
> heuristics are trying to achieve? What are the sorts of species
> divisions found in taxonomies? This is an area where more would be
> helpful - and where a shorter introduction might offer space.

> 14-40. Whereas just about every 16S survey investigation involves an
> OTU-based alpha diversity estimate, only a few involve PD: unweighted
> PD has been applied to some 16S survey data (Lozupone and Knight 2007;
> Costello et al. 2009) and to metagenomic reads in a set of marker
> genes (Kembel et al. 2011) - This is an odd pair of sentences: is the
> L&K the only study that used unweighted PD? If so, in what context?

I can't quite understand this suggestion. All three papers listed used
unweighted PD.

@@

> 14-46. Check this sentence - perhaps start with "While are commonly
> used, abundance weighted PD measures are not.

I have updated this sentence based on your suggestion, thanks.

@@

> 15-30. Perhaps "In contrast to measures of alpha diversity, measures
> of beta diversity do make use of phylogenetics"  This is actually an
> interesting contrast. Is this just historical contingency, or is there
> a logical reason why the two areas disagree on the use of phylogenetic
> information?

I think that this refers to 14-30, and I've simply removed the sentence.
With regards to your question, I don't know. It seems surprising to me.

@@

> 15-xx. This seems a strong point of contact between metazoan
> phylogenetics and the microbiome folks. UniFrac is clearly described
> on its own terms, except perhaps where I would have written "with the
> fraction of edges connecting samples that are common to two samples"
> rather than "edges that connect between samples" (which doesn't
> strictly make sense). I believe there is the same measure of
> beta-diversity in the community-phylogenetics literature? What is the
> closest analogue?

I don't follow your "with the fraction of edges connecting samples that are common to two samples" correction.
There is some overlap with the PD literature as now described and cited.

### 2014-04-11 sbreview.tex
Done for the day.

Despite the enthusiasm with which microbial ecologists have accepted between-community comparison using phylogenetics (next section), phylogenetic alpha diversity seems under-developed.
Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are among the most common, abundance weighted phylogenetic diversity measures are not used in human microbiome studies.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

Faith2009-gs

>>>>

Phylogenetic diversity (PD) measures use the structure and branch lengths of a phylogenetic tree to measure the diversity of a sample (Fig.~1).
The original (unweighted) PD of a set of taxa is simply the total of the lengths of the branches connecting taxa in the set \citet{faith1992conservation}.
It quantifies the ``amount of evolution'' contained in the evolutionary history of those taxa.
Whereas just about every 16S survey investigation involves an OTU-based alpha diversity estimate, only a few involve PD: unweighted PD has been applied to some 16S survey data \citep{lozupone2007global,costello2009bacterial} and to metagenomic reads in a set of marker genes \citep{kembel2011phylogenetic}.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
\citet{mccoy2013abundance} have recently shown that partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \cite{Faith2009-gs}, who point out related measures in Faith's earlier work.
